
# Development and External Validation of a Multi-State Clinical Prediction Model for Chronic Kidney Disease Patients Progressing onto Renal Replacement Therapy and Death - Supplementary Material{#chap-dev-paper-supp}
`r fb(thesis="\\chaptermark{Development and Validation of MSCPM - Supp}")`
<script src="https://hypothes.is/embed.js" async></script>

## Statistical Analysis

### Development

Data was recorded in a time-updated manner, however all variables were measured at baseline to emulate the real-world application of the model (i.e. future prediction of states and not covariates).  As well as those included in `r fb(gitbook="\\@ref(chap-dev-paper)", thesis="\\@ref(chap-dev-paper)", single="the main text")`, ethnicity was assessed in the populations, but as most patients were white, it was omitted as a potential predictor from the models. 

 log(Calendar Time) was included as a covariate to adjust for secular trends in treatment preferences [@bhatnagar_epidemiology_2015]. Calendar Time was defined as length of time between start date and  1st January 2019.

Intermediate states (RRT or modality) were considered to be medically transformative, and so a semi-markov (clock reset) method for analysis was considered to be well justified [@meira-machado_multi-state_2009]. Each transition was modelled under a proportional hazards assumption using the Royston-Parmar technique [@royston_flexible_2002] to estimate coefficients for each covariate and a restricted cubic spline (on the log-time scale) for the baseline cumulative hazard. The cumulative hazards for each transition can be combined to produce estimates for the probability of a patient being in any state at any time [@putter_tutorial_2007].

For variable selection, we stacked the imputed datasets together to create a larger, pseudo-population [@wood_how_2008] and performed backwards-forwards selection based on minimising the AIC at each step. This was repeated for each transition and for different numbers of evenly spaced knots in modelling the form of the baseline hazard, K={0,1,2,3,4,5}. This allowed for different transitions to use different sets of variables and numbers of knots in the final model. Some combinations of variables resulted in models that were intractable and so these models were excluded. Once a set of variables were chosen, the R-P model was applied to each imputed dataset individually and the resulting coefficients and cubic spline parameters were aggregated across imputations using Rubin's Rules [@rubin_multiple_1984]. This gave a model fully defined by smooth cubic splines representing the cumulative cause-specific hazard and individualised proportional hazards for each transition.

All missing data were assumed to be missing at random and so were multiply imputed using chained equations with the Nelson-Aalen estimators for each relevant transition as predictors [@white_imputing_2009]. Some variables (smoking status and histories of COPD, LD and ST)  were present in the SKS (development) dataset, but were completely missing in the SERPR extract (validation) and so these were multiply imputed from the development dataset [@janssen_dealing_2009].



### Validation



For validation purposes, we consider Death and Death after RRT/HD/PD to be distinct states meaning that for the Three-State model, we have $K=4$ pathways a patient can take and for the Five-State model, we have $K=7$. To compare across models, we combined states together to collapse down to simpler versions. We collapsed the Three-State model to a two-state structure by combining the CKD and RRT states into an Alive state. We collapsed the Five-State model to a three-state structure by combining the HD, PD and Tx into an RRT state and then further down to a two-state structure as with the Three-State model. We will report performance measures at 360 days (approx. 1-year), 720 days (approx. 2-years) and 1800 days (approx. 5-years). As well as presenting the performance measures over time.

`r fb("The performance metrics were chosen from those defined in chapter \\ref@(chap-performance-metrics)","The performance metrics were chosen from those defined in chapter \\ref@(chap-performance-metrics)","")`

The overall accuracy of each model was assessed using the MSM adjusted Brier Score`r fb(single=" [@chap-performance-metrics]")`, which is a proper score function [@gneiting_strictly_2007] assigning 0 to a non-informative model and 1 to a perfect model, with negative numbers implying the model performs worse than assuming every patient's state predictions are the same as the overall prevalence within the population.

The discrimination of each model was assessed using the MSM extension to the c-statistic [@calster_extending_2012-1`r fb(single=";@chap-performance-metrics")`]. The c-statistic is a score between 0 and 1 with higher scores suggesting a better model and a c-statistic of 0.5 suggesting the model performs no better than a non-informative model.

The calibration of each model was assessed using MSM multinomial logistic regression (MLR)  [@hoorde_assessing_2014`r fb(single=";@chap-performance-metrics")`] which extends the logistic regression to three or more mutually exclusive outcomes [@riley_prognosis_2019]. This produces an intercept vector of length $K-1$ and a Slope-matrix of dimension $(K-1) \times (K-1)$. As with the traditional calibration intercept for a well performing model, the MLR intercept values should all be as close to 0 as possible.  The traditional calibration slope should be as close to 1 as possible and so the multi-state extension of the slope, the Slope-matrix should be as close to the identity matrix ($I$) as possible.

## Model Results

### Two State Model


Table \@ref(tab:PH-Two) shows the proportional hazard ratios for the transitions in the Two-State Model. Older patients have a higher hazard towards death, low adn decreasing eGFR increased hazard as did a history of diabetes. Patients with a primary renal diagnosis included in the ERA-EDTA [@venkat-raman_new_2012] definition of Systemic diseases affecting the kidney had the highest likelihood of death.

```{r PH-Two,echo=F}
print_table_PH("data/Dev_Paper_Hazard_Ratios.csv",2,lscape=F)
```

Equation \@ref(eq:CH-Two-16) below shows the baseline cumulative hazard functions for the transition from Alive to Dead in the Two-State Model.

`r print_Cum_Haz_eq("data/Dev_Paper_Baseline_Cumulative_Hazard.csv",2)`

Table \@ref(tab:IV-Two) shows the results from the internal validation in the Two-State Model. Calibration Intercept is close to 0, implying the model is well calibrated overall with a high c-statistic and Brier Score. Calibration Slope above 1 implies that the model under-estimates outcomes.


```{r IV-Two,echo=F}
res <- list_table_Valid("data/Dev_Paper_Validation.csv",2,F)
res[[1]]
```

Table \@ref(tab:EV-Two) shows the results from the external validation in the Two-State Model, which shows similar results to the internal validation with slightly impaired perfomance, which is to be expected in an external validation.


```{r EV-Two,echo=F}
res <- list_table_Valid("data/Dev_Paper_Validation.csv",2,T)
res[[1]]
```

### Three State Model






The equations \@ref(eq:CH-Three-16), \@ref(eq:CH-Three-15) and \@ref(eq:CH-Three-56) shows the baseline cumulative hazard functions for the transition from CKD to Dead, CKD to RRT and RRT to Dead, respectively in the Three-State Model.

`r print_Cum_Haz_eq("data/Dev_Paper_Baseline_Cumulative_Hazard.csv",3)`

Validation results for the Three-State Model can be found in `r fb(gitbook="Chapter \\@ref(chap-dev-paper)", thesis="Chapter \\@ref(chap-dev-paper)", single="the main text")`.





### Five State Model

Table \@ref(tab:PH-Five) shows the proportional hazard ratios for the transitions in the Five-State Model.

```{r PH-Five,echo=F}
print_table_PH("data/Dev_Paper_Hazard_Ratios.csv",5,lscape=T)
```

The equations \@ref(eq:CH-Five-12), \@ref(eq:CH-Five-13), \@ref(eq:CH-Five-14) and \@ref(eq:CH-Five-16) show the baseline cumulative hazard functions from the CKD state to HD, PD, Tx and Dead, respectively. Equation \@ref(eq:CH-Five-26) shows the baseline cumulative hazard function from HD to Dead and Equation \@ref(eq:CH-Five-36) shows the baseline cumulative hazard function from PD to Dead.

`r print_Cum_Haz_eq("data/Dev_Paper_Baseline_Cumulative_Hazard.csv",5)`

Table \@ref(tab:IV-Five) shows the results from the internal validation in the Five-State Model. The calibration slope results are shown in a seperate table for both the internal and external validation.


```{r IV-Five,echo=F}
res_IV <- list_table_Valid("data/Dev_Paper_Validation.csv",5,F)
res_IV[[1]]
```


Table \@ref(tab:EV-Five) shows the results from the external validation in the Five-State Model.

```{r EV-Five,echo=F}
res_EV <- list_table_Valid("data/Dev_Paper_Validation.csv",5,T)
res_EV[[1]]
```

Table \@ref(tab:Five-Valid-Slope) shows the calibration slopes for the model in the internal and external datasets in both the < 60 eGFR and <30 eGFR sub-populations.


```{r Five-Valid-Slope, echo=F}
full_join(res_IV[[2]],res_EV[[2]],
          by="Time") %>%
  to_kable(caption="Calibration Slope results for both the External and Internal Validation for the Five-State Model",align="rr",row_names="Time",numeric_cols=2:5,col_widths=rep("20em",5))
```




